Visualizing Contingency Tables

### 2006-04-20 trunk/vcd/paper/hsc.Rnw
first fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@521 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. They have been
extended over the last two decades, and implementations exist in many
statistical environments.

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms \cite{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

This chapter will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Section~\ref{sec:concl} concludes the chapter.

@@

Although far from optimal, contingency tables are frequently visualized using
grouped bar plots (see Fig.~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Fig.~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If the variables can be
separated into explanatory and dependent variable(s), the explanatory variable(s)
should be be used first for splitting to display the conditional distribution of
the dependent variable(s) given the explanatory variable(s). 
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary interest.

@@

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

In Fig.~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

For association plots, residual-based shadings are redundant since all relevant
information is already contained in the plot by construction. Nevertheless, using one of the
shadings discussed above will support the analysis process and is therefore recommended.
For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
on the one hand facilitates the discrimination between
positive and negative residuals, and on the other, more importantly,
allows an easier comparison of the tiles' sizes. Thus, the shading
supports the detection of the pattern.

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Fig.~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
time using the maximum test statistic and its 10\% and 1% critical values at cut-off points.
The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

@@

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is 
to draw a doubledecker plot nesting gender in
department and highlighting admission in the tiles (see Fig.~\ref{fig:doubleucb}).
From the widths of the tiles it can be seen that an
unequal number of students apply for the six
departments, and from the highlighting we see that the admission rates differ as
well among the departments (roughly speaking, they decrease from A to F). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

A first approach is to use pairs plots for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others, etc.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

>>>>

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. Especially
mosaic and sieve plots have been extended over the last two decades, 
and implementations exist in many statistical environments.

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001} or spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}.
%% or spinograms \cite{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Implementations of association and sieve plots can only be found in
\proglang{R} and \proglang{SAS} (in the latter, these plots are available
for two-way tables only).
Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Figure~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network
(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.

@@

This chapter will give an overview of the state of the art 
of mosaic and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. The techniques are illustrated using three- and four-way tables.
Section~\ref{sec:concl} concludes the chapter.

@@

Although far from optimal, contingency tables are frequently visualized using
grouped bar plots (see Figure~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Figure~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Figure~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.

@@

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. 
If we cross-tabulate each tile to fill it with a number of squares equal to
the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
(see Figure~\ref{fig:sievefig}). It implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the density of the rectangles is marked along the secondary
diagonal, indicating a negative association of 
the two variables. This gives evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If one variable is
explanatory, it should be be used first for splitting: thus, the display shows
the conditional distribution of the dependent variable given the explanatory one.
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models are of primary interest.

@@

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminance better 
contrasts can be achieved.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

In Figure~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

For association plots, residual-based shadings are redundant since all relevant
information is already contained in the plot by construction. Nevertheless, using one of the
shadings discussed above will support the analysis process and is therefore recommended.
For example, applying the Friendly shading in Figure~\ref{fig:colorassoc} 
on the one hand facilitates the discrimination between
positive and negative residuals, and on the other, more importantly,
allows an easier comparison of the tiles' sizes. Thus, the shading
supports the detection of the pattern.

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Figure~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
time using the maximum test statistic and its 10\% and 
1\% critical values as cut-off points.
The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

@@

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that shall be ``explained'' by the others. 
A natural way of representing such a three-way table is 
to draw a doubledecker plot nesting gender in
department and highlighting admission in the tiles (see Figure~\ref{fig:doubleucb}).
From the widths of the tiles it can be seen that an
unequal number of students apply for the six
departments, and from the highlighting we see that the admission rates differ as
well among the departments (roughly speaking, they decrease from A to F). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

A first approach is to use pairs plots for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?
